Spatial and temporal variability in economic availability of residual biomass

### 2021-12-01 analysis/newms_elsevier.Rmd
Offloaded to Jay & Alex

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990--2015)
author:
  - name: Glen W. Armstrong
    email: gwa@ualberta.ca
    affiliation: rr
    footnote: 1
  - name: Jay A. Anderson
    email: jaya@ualberta.ca
    affiliation: rees
  - name: Alex W. McPhee
    email: awmcphee@ualberta.ca
    affiliation: rees
address:
  - code: rr
    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
  - code: rees
    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

@@

journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
bibliography: historicalResidueSupply.bib
linenumbers: true
numbersections: true
csl: elsevier-vancouver.csl
output: rticles::elsevier_article
---

@@

1. Availability of data
2. Temporal variability
3. suitability for mid-level analysis
4. ultra-cool graphics

* renewable fuels better than fossil fuels
* forest harvest residue just piled and burned
* forest road networks in Alberta radiate from location
* terrain is difficult
* simple circle or doughnut models of limited utility

Humans have been using biofuels since we learned to control fire a
million or so years ago.  In developed economies, biofuels were
largely displaced by coal during the industrial revolution, and by
petroleum starting in the mid-nineteenth century.  In recent times,
biofuels are an increasingly important source of energy because of
their potential to reduce dependency on fossil fuels and reduce
greenhouse gas emissions.  Direct combustion of biomass is still an
important source of energy.  In Europe, in particular, many
electricity plants have been converted to use wood pellets instead of
coal as the primary energy source.  Liquid biofuels (*e.g.* ethanol
and biodiesel) have been produced in commercial quantities from food
crops starting in the 1980s.  These are called first generation
biofuels.  A major difficulty with these first generation biofuels is
that they compete with food and that they may lead to land use change.

Second generation biofuels rely on processes that convert
lignocellulosic materials (such a corn stover and wood) to liquid
biofuels.  An advantage of this process is that the feedstock can be
residue from food or forest products production.  Second generation
biofuel production is still largely in the pilot phase.  An
interesting example using forest residues is the Silva Green Fuel
renewable crude oil plant in Tofte, Norway which is expected to begin
operations in 2021 @Lane2018.

Wood pellets are a well-established biofuel.  It is projected that
Canada will produce 3.8 million Mg of wood pellets in 2021, of which
87% will be exported, largely to the United Kingdom, Japan, and Europe
@Watters2021.  Wood pellets are replacing coal as the primary energy
source in electricity generating plants, particularly in Europe
@Sterman2018.  Most of the wood pellet production in Canada comes from
mill waste.  There is interest in expanding the source to the forest
by collecting the residue left on cutblocks after harvest.  Forest
residues are generated as by-products of conventional forest
harvesting operations, and include the tree tops, branches and
non-merchantable logs. These residues are typically left behind in
slash piles, which are burned during winter to mitigate against the
risk of wildfires. Canada has a forestry sector that produces
approximately 20 million oven-dry metric tons of forest residues
annually @Dymond2010.  For the remainder of this paper we will express
biomass using SI units, and all quantities of biomass will be
expressed on an oven-dry basis: 1 oven-dry metric ton is equivalent to
1 Mg.

<!-- Clean fuel standard https://gazette.gc.ca/rp-pr/p1/2020/2020-12-19/html/reg2-eng.html  -->

<!-- Despite the potential to produce biofuels from forest residues, such use -->
<!-- is limited to a few highly subsidized pilot projects. Several -->
<!-- operational and economic challenges hinder the utilization of forest -->
<!-- residues. These challenges are related to: (1) capital investments in -->
<!-- biorefineries, (2) geographical distance between feedstock and markets, -->
<!-- and (3) concerns surrounding feedstock availability @Shabani2013. The purpose of our study is to explore concerns surrounding this -->
<!-- third challenge -- feedstock availability -- by evaluating annual -->
<!-- variability in commercial stemwood harvest in the province of Alberta, -->
<!-- Canada. -->

A number of studies have shown that collection and transportation
costs are arguably the biggest constraint to widespread use of forest
residues in bioenergy systems
[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
for a bioenergy plant largely depend on commercial harvesting operations
and the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products has a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in
forest harvesting activities
@deAvillez2014 @Spelter2009. This relationship between volatile
harvest volumes and subsequent residue variability is explored in detail by
Niquidet and Friesen @Niquidet2014, who model forest residue
supply in Alberta as a  function of lumber prices. Given that
biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
costs @Stephen2010, reliable feedstock supplies
and costs over the life of the bioenergy plant are important factors to
consider when deciding where to locate the bioenergy plant.

Though past studies regarding bioenergy plant locations have greatly
improved our understanding of feedstock availability and constraints,
one key area that has largely been omitted is considerations of
feedstock supply variability over time. Commercial stemwood harvests
vary year-to-year, largely based on market conditions for forest
products. Since a forest-residue based bioenergy plant is dependent on
stemwood harvesting for feedstock, any variability in harvesting
directly influences the volume of residues that would be available.
Therefore, considering the variability associated with feedstocks in
different locations could be important in choosing the optimal site for
locating a bioenergy plant

Although not related to a bioenergy plant siting decision, some studies
have considered variable feedstock supplies when investigating whether
and when to expand the production of bioenergy from burning residues
 @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
studies have explored the context of variable supplies when making
decisions about capital investments in forest residue-based
biorefineries  @Cambero2015, @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment
decisions related to: investing in a residue preprocessing facility
(Chen and Fay 2011) @Chen2012, types of storage infrastructure for
bioenergy plant feedstocks (Papapostolou and Kondili 2011)
@Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities  @Cambero2015.

There has been a number of studies conducted regarding siting decisions
for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
a review of these studies -- however, we are aware of no study that
considered stemwood harvest variability and its subsequent effect on
forest residue variability. Therefore, our study contributes to the
literature by considering variability in forest residue feedstock
supplies. More specifically, we analyze how this variability affects the
cost of delivering residue, and then use these cost estimates to inform
the decision on where to locate a bioenergy plant.

We consider co-locating the bioenergy adjacent to either a sawmill
or a pulpmill. Co-locating with existing facilities and making use of
established infrastructure and transportation systems can reduce many
of the costs associated with establishing a new bioenergy plant
@Browne2012 @Saddler2012 @Benjamin2009 @Towers2007 The decision to
co-locate a bioenergy plant near a sawmill or pulpmill should be informed
by how stable the respective residue supplies are. The facility with a
more stable stemwood harvest would generate a more stable supply of
forest residues that could be collected from the nearby harvested
areas.

The objectives of our study are ...... Our approach incorporates
variable feedstock availability into decision making, and our results
provide important information for assessing the potential for a forest
residue-based biofuel industry in Alberta.

The remainder of the paper is as follows: The next section describes

The methods used for this study are similar to that used by
FPInnovations for their biomass availability estimations for timber
supply areas in British Columbia (see Ref. @Blackburn2020 for an
example).  Like us, FPInnovations develops supply curves
(cost-availability curves based on a road network and historical
harvests).  However, their take on temporal variability is limited to
examining 2 5-year periods. As well, they limit their analyses to the
boundaries of timber supply areas.  We do not limit ourselves with
arbitrary boundaries.  Supply cost curves.

Our analysis also shares some similarities with work published by
Yemshanov *et al.* @Yemshanov2014. National level based on costs to
cogeneration facilities. Potential harvest locations.   Again our interest included
temporal variability. actual harvest locations.
<!-- ....... -->

<!-- **OTHER POTENTIALLY USEFUL TIDBITS..** -->

<!-- The need to go outside its normal delivery zone in bad years is similar -->
<!-- to a scenario in the agriculture residue feedstock supply literature, -->
<!-- called the the "derisked" scenario, whereby a buffer area outside the -->
<!-- normal supply zone is contracted to provide residues during years when -->
<!-- residue supply is low @Golecha2016. -->

<!-- Modern biorefineries can expect to breakeven financially with -->
<!-- delivered biomass costs of approximately \$80 (US\$60) t\[oven -->
<!-- dried\]^-1^ (Steeper 2021) @Steeper2021 . Studies have shown that -->
<!-- forest policy reforms could reduce delivered forest residue costs by -->
<!-- incenting plantations of fast-growing tree species on lands close to -->
<!-- pulpmill locations (Shooshtarian et al. 2021 @Shooshtarian2021, -->
<!-- Anderson et al. 2012 @Anderson2012). Other important factors will be -->
<!-- advances in biorefinery technology and increases in biofuel prices. In -->
<!-- Canada, governments impact biofuel prices by setting renewable fuel -->
<!-- mandates and paying subsidies for biofuel production (Campbell et al., -->
<!-- 2016). The continuation of these mandates and subsidies will likely be -->
<!-- an important factor for the emergence of a second-generation -->
<!-- ("advanced") biofuel sector. Also, the Government of Canada\'s Clean -->
<!-- Fuel Standard, which is proposed to come into force in 2022 -->
<!-- (Government of Canada, 2020), is expected to improve the economics of -->
<!-- advanced biofuel production.   -->`

@@

We describe here our study area, data sources, and method.

@@

The southeastern part of the province is largely privately owned
agricultural land, but the northern and western parts are largely
publicly owned forested land.  There is a substantial area of
privately owned agricultural land in the Peace River Country of
northwestern Alberta.  There are also substantial areas in National
Parks along the Rocky Mountains and in the far north of the province.
The province calls the area of mostly publicly owned forested land
(excluding those National Parks) as the Green Area, and the area of
mostly privately owned agricultural land as the White Area.  About 90%
of the harvested timber volume in Alberta comes from the Green Area.

For administrative purposes, the Green Area is divided into Forest
Management Units (FMUs).  The province also enters into long term
agreements with forest products companies giving the companies the
rights to timber harvest and the responsibility to manage the forest
sustainably.  These agreements are called Forest Management
Agreements, and the areas under agreement are called Forest Management
Agreement Areas (FMAAs).  For the most part, the boundaries of the
FMAAs are now coincident with FMU boundaries, with the exception of
the Alberta-Pacific FMAA in eastern Alberta comprises 12 FMUs.  Maps
of the FMUs @AlbertaFMU and the FMAAs @AlbertaFMA can be found on the
Government of Alberta website.  The FMU and FMAA boundaries are
important to this study as they are used to look up average piece size
and utilization standards.  We discuss piece sizes and utilization
standards later.

Our focus for this study is the forest near Hinton, Edson, and
Whitecourt in west-central Alberta.  These towns were chosen as they
host several forest products companies and are located near several
active FMAAs.  We assume that a residual forest biomass processing
plant would be located near an existing forest products mill as the
forest road network has been developed to supply those mills. Locating
a forest residue processing plant near a forest products mill would
also allow for convenient access to mill residues.

@@

Altalis manages much of the spatial data created by or for the
Government of Alberta.  We use several coverages from their base
features database product @Altalis2021: the provincial, Green/White
area, forest management unit (FMU), forest management agreement area
(FMAA), Alberta Township Survey System township, and town boundaries.

The Alberta Township Survey System (ATS) is particularly important to
this study as all the data used for this study are aggregated or
disaggregated to the township level.  ATS provides the basis for legal
land descriptions in Alberta.  The typical township is roughly square
(9.778 km by 9.716 km, 9 500 ha) and is indexed by township (1--126),
range (1--30), and meridian (4--6).  The townships immediately
east of Alberta's western boundary, and immediately east of the 5^th^
and 6^th^ meridians (corresponding to 114°W and 118°W longitude,
respectively) are partial townships and therefore deviate from the
roughly square shape.  ATS is described in detail by the Alberta Land
Surveyor's Association @ALSA2021.  There are 7 237 townships in
Alberta, allowing for spatial resolution appropriate for regional or
provincial level analyses.

@@

Alberta Geospatial Services is operated by the Alberta Ministry of
Environment and Parks @Alberta2021a.  We obtained road access
coverages from here.  A road access coverage is also available from
Altalis but it does not have as detailed information on road class.

@@

The National Forest Information System (NFIS) provides a set
of spatial data layers representing forest properties at a 250 m
resolution for all of Canada @Beaudoin2017.  We used the merchantable
volume, stand age, and species composition attributes from these
data layers.

Another product of NFIS is a set of spatial datasets derived from
satellite data @NFIS2021a.  The dataset used for this study is the
harvest year data set for the years 1985-2015.  The dataset is
described in detail by Hermosilla *et al.*  @Hermosilla2016.  It
consists of a raster for Canada at 30 m resolution indicating the year
of harvest for each cell.

@@

Ung *et al.*  @Ung2013 describe the tree taper models 
developed for Canada.  A valuable contribution related to this is the WWW-based
wood volume calculator that uses those models @NRCan2015.  We use the
taper models developed by Ung *et al.* to calculate merchantable
volumes and the amount of residual biomass available from cutblocks in
each of the townships in each year.

@@

The national forestry database @CCFM2021 was used to summarize the 
timber harvest volume and harvest area from
Alberta for the years 1990-2015. This information is used to adjust the volumes
obtained from the 250 m resolution inventory to better reflect average harvest
volumes observed in Alberta.

@@

The Alberta government requires FMA holders to develop a forest management
plan for the FMAA.
The forest management plans were used to get information on utilization 
standards and average piece size of harvested trees for the FMAA @Alberta2021.

@@

The boundaries of Canada's terrestrial ecozones can be found on
Agriculture and Agri-Food Canada's website @Canada2017.  Ecozone is an
input to the individualt tree biomass calculator @NFIS2021b.

@@

The primary objective of this paper was to quantify the spatial and
temporal variability of the costs associated with recovering the
residual biomass left on forest cutblocks following timber harvest.  A
driving assumption for this model was that residue recovery is
opportunistic.  Loggers make the decision about what and where to
harvest, and the biomass recovery occurs at some time following timber
harvest.  The decision to harvest is made independently of any
knowledge of the value of residue as biomass.  The biomass gleaners
provide a service to the loggers as the residue would be piled and 
burned if it was not recovered as biomass.

Every township will have associated with it a cycle time (and cost)
required to transport harvest residue from the township to each of our
3 plant locations, a measure of residual biomass for each of the 26
years in the study period. For each plant location and year, the model
looks at all the townships in the province that have a positive
residual volume, and begins the creation of cumulative residual
biomass curves starting with the lowest cycle time first.  These lists
of cycle time and cumulative biomass provide the data for the marginal
cycle time and cost curves.

@@

The provincial road network was sorted into four classes: primary
highways, other paved roads, two-lane gravel roads, and one-lane
gravel or dirt roads.  Loaded and unloaded travel speeds were assumed
for each road class, according to Table 1. These travel speeds are
based on those assumed by FPInnovations for the Dawson Creek Timber
Supply Area in the neighbouring province of British Columbia
@Blackburn2020.

The harmonic mean of loaded and unloaded speeds was calculated as 
$$
\bar{s_h}=\left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1} 
$$ 
where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.  Use
of harmonic mean allows us to calculate the correct cycle time: the
time required to travel from the mill, to the collection site at the
cutblock, and back to the mill.  This provides the basis for our
estimates of haul cost.

Figure 3 illustrates the road network in Alberta and estimates of
cycle time based on the average of loaded and unloaded speeds
from Table 1. The road network in the White Area of the province is
very dense and largely arranged as a grid related to the Alberta
Township Survey System.  Most of the roads in the White Area are
public.  The road network in the Green Area is less dense and
many of the roads are built and maintained by resource industries
(*e.g.* forestry and petroleum & natural gas).  The pattern of the
road network is not as regular as in the White Area.

[ Figure 3 about here]

One-way haul times based on the harmonic mean travel speed were
generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
created by Raffler @Raffler2018, specifically its iso-area as
interpolation (from point) algorithm.  In our case, we used the
algorithm to determine the minimum one-way travel time (using the
harmonic mean of loaded and unloaded speeds) along the road
network from each of our centres to each cell of a provincial level
raster at a 200 m resolution.  Off-road travel was assumed to be at 5
km/h.  The underlying algorithm for travel time minimization is that
of Dijkstra @Dijkstra1959.  Cycle times were calculated by multiplying
the one-way haul time by 2, and adding 1 hour for loading and unloading

@@

[ Figure 2 about here ]

@@

All measures of biomass used in this study are reported in SI units on
a dry matter basis:  1 Mg is equivalent to 1 oven dry metric 
ton.

>>>>

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990--2015)
author:
  - name: Glen W. Armstrong
    email: gwa@ualberta.ca
    affiliation: rr
    footnote: 1
  - name: Jay A. Anderson
    email: jaya@ualberta.ca
    affiliation: rees
  - name: Alex McPhee
    email: awmcphee@ualberta.ca
    affiliation: rees
address:
  - code: rr
    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
  - code: rees
    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

@@

journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
bibliography: historicalResidueSupply.bib
linenumbers: true
numbersections: true
csl: elsevier-vancouver.csl

output: rticles::elsevier_article
---

@@

Humans have been using biofuels since we learned to control fire a
million or so years ago.  In developed economies, biofuels were
largely displaced by coal during the industrial revolution, and by
petroleum starting in the mid-nineteenth century.  In recent times,
biofuels are an increasingly important source of energy because of
their potential to reduce dependency on fossil fuels and reduce
greenhouse gas emissions.  Direct combustion of biomass is still an
important source of energy.  In Europe, in particular, many
electricity plants have been converted to use wood pellets instead of
coal as the primary energy source.  Liquid biofuels (*e.g.* ethanol
and biodiesel) have been produced in commercial quantities from food
crops starting in the 1980s.  These are called first generation
biofuels.  A major difficulty with these first generation biofuels is
that they compete with food and that their use  may lead to land use change.

Second generation biofuels rely on processes that convert
lignocellulosic materials (such a corn stover and wood) to liquid
biofuels.  An advantage of this process is that the feedstock can be
residue from food or forest products production.  Second generation
biofuel production is still largely in a pilot or demonstration phase.  An
interesting example using forest residues is the Silva Green Fuel
renewable crude oil plant in Tofte, Norway which is expected to begin
operations in 2021 @Lane2018.

Wood pellets are a well-established biofuel.  It is projected that
Canada will produce 3.8 million Mg of wood pellets in 2021, of which
87% will be exported, largely to the United Kingdom, Japan, and Europe
@Watters2021.  Wood pellets are replacing coal as the primary energy
source in electricity generating plants, particularly in Europe
@Sterman2018.  Most of the wood pellet production in Canada comes from
mill residues.  There is interest in expanding the source to the forest
by collecting the residue left on cutblocks after harvest: this
practice is well-established in Europe.  Forest
residues are generated as by-products of conventional forest
harvesting operations, and include the tree tops, branches and
non-merchantable logs. These residues are typically left behind in
slash piles, which are burned during winter to mitigate against the
risk of wildfires. Canada has a forestry sector that produces
approximately 20 million oven-dry metric tons of forest residues
annually @Dymond2010.  For the remainder of this paper we will express
biomass using SI units, and all quantities of biomass will be
expressed on an oven-dry basis: 1 oven-dry metric ton is equivalent to
1 Mg.

~~Clean fuel standard
https://gazette.gc.ca/rp-pr/p1/2020/2020-12-19/html/reg2-eng.html
Despite the potential to produce biofuels from forest residues, such
use is limited to a few highly subsidized pilot projects. Several
operational and economic challenges hinder the utilization of forest
residues. These challenges are related to: (1) capital investments in
biorefineries, (2) geographical distance between feedstock and
markets, and (3) concerns surrounding feedstock availability
@Shabani2013. The purpose of our study is to explore concerns
surrounding this third challenge -- feedstock availability -- by
evaluating annual variability in commercial stemwood harvest in the
province of Alberta, Canada.~~

A number of studies have shown that collection and transportation
costs may be the biggest impediment to widespread use of forest
residues in bioenergy systemsX
[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
for a bioenergy plant largely depends on commercial harvesting operations
and the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products has a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in
forest harvesting activities
@deAvillez2014 @Spelter2009. This relationship between volatile
harvest volumes and subsequent residue variability is explored in detail by
Niquidet and Friesen @Niquidet2014, who model forest residue
supply in Alberta as a  function of lumber prices. Given that
biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
costs @Stephen2010, reliable feedstock supplies
and costs over the life of the bioenergy plant are important factors to
consider when deciding where to locate the bioenergy plant.

Past studies regarding bioenergy plant locations have
improved our understanding of feedstock availability and constraint.
One key area that has largely been omitted is considerations of
feedstock supply variability over time. Commercial stemwood harvests
vary year-to-year, largely based on market conditions for forest
products. Since a forest-residue based bioenergy plant is dependent on
stemwood harvesting for feedstock, any variability in harvesting
directly influences the volume of residues that would be available.
Therefore, considering the variability associated with feedstocks in
different locations could be important in choosing the optimal site for
locating a bioenergy plant

**We say above that consideration of variable feedstock supplies has
been omiXtted, and in the paragraph below that a bunch of people have
consider it.  Do those studies actually consider variability?**

~~Some studies
have considered variable feedstock supplies when investigating whether
and when to expand the production of bioenergy from burning residues
 @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
studies have explored the context of variable supplies when making
decisions about capital investments in forest residue-based
biorefineries  @Cambero2015, @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment
decisions related to: investing in a residue preprocessing facility
(Chen and Fay 2011) @Chen2012, types of storage infrastructure for
bioenergy plant feedstocks 
@Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities  @Cambero2015.~~

~~There has been a number of studies conducted regarding siting decisions
for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
a review of these studies -- however,~~

We are aware of no study that
considers historical variability of stemwood harvest and its subsequent effect on
forest residue availability. Our study contributes to the
literature by considering variability in forest residue feedstock
supplies.

More specifically, we analyze how this variability affects the
cost of delivering residue, and then use these cost estimates to inform
the decision on where to locate a bioenergy plant.

We consider co-locating the bioenergy adjacent to existing forest
products manufacturing facilties. Co-locating with existing facilities
and making use of established infrastructure and transportation
systems can reduce many of the costs associated with establishing a
new bioenergy plant @Browne2012 @Saddler2012 @Benjamin2009 @Towers2007
(**I don't have the references for Browne2012, Saddler2012,
Benjamin2009, or Towers2007**).  Bioenergy plant location should be
informed by how stable the respective residue supplies are. More
stable stemwood harvest would generate a more stable supply of forest
residues that could be collected from the nearby harvested areas.

We use information on the location and year of timber harvest in the
province of Alberta, Canada to quantify the spatial and temporal
variability of timber harvest, and relate that information to spatial
and temportal variability of production of timber harvest residues.
We develop transportation cost estimates to the timber harvest areas
based on cycle time estimates developed from the road network in the
province. From this information we develop historical techno-economic
supply curves [@Galik2009,@Yemshanov2014,@Blackburn2020, @Swinton2021]
for forest residue biomass for each of 26 years in our study period
(1990--2015) for 3 potential biofuel plant locations.

~~We use information The objectives of our study are ...... Our approach
incorporates variable feedstock availability into decision making, and
our results provide important information for assessing the potential
for a forest residue-based biofuel industry in Alberta.~~

The methods used for this study are similar to that used by
FPInnovations for their biomass availability estimations for timber
supply areas in British Columbia (see Ref. @Blackburn2020 for an
example).  Like us, FPInnovations develops supply curves
(cost-availability curves) based on a road network and historical
harvests.  However, their take on temporal variability is limited to
examining 2 5-year periods. As well, they limit their analyses to the
boundaries of timber supply areas.  We do not limit ourselves with
arbitrary boundaries.

Our analysis also shares similarities with work published by
Yemshanov *et al.* @Yemshanov2014. They devolop supply curves at
national and regional levels based on a national forest inventory and
the location of cogeneration facilities associated with forest
products manufacturing facilities.

The main contribution of this paper is the consideration of multiple
years to develop estimates of the temporal variability in forest
residue supply.

~~OTHER POTENTIALLY USEFUL TIDBITS.. I left these here as a reminder.~~

~~The need to go outside its normal delivery zone in bad years is similar
to a scenario in the agriculture residue feedstock supply literature,
called the the "derisked" scenario, whereby a buffer area outside the
normal supply zone is contracted to provide residues during years when
residue supply is low @Golecha2016.~~

~~Modern biorefineries can expect to breakeven financially with
delivered biomass costs of approximately \$80 (US\$60) Mg (Steeper
2021) @Steeper2021 . Studies have shown that forest policy reforms
could reduce delivered forest residue costs by incenting plantations
of fast-growing tree species on lands close to pulpmill locations
(Shooshtarian et al. 2021 @Shooshtarian2021, Anderson et al. 2012
@Anderson2012). Other important factors will be advances in
biorefinery technology and increases in biofuel prices. In Canada,
governments impact biofuel prices by setting renewable fuel mandates
and paying subsidies for biofuel production (Campbell et al.,
2016). The continuation of these mandates and subsidies will likely be
an important factor for the emergence of a second-generation
("advanced") biofuel sector. Also, the Government of Canada\'s Clean
Fuel Standard, which is proposed to come into force in 2022
(Government of Canada, 2020), is expected to improve the economics of
advanced biofuel production.~~

@@

We describe here our study area, data sources, and methods.

@@

The southeastern part of the province is largely privately owned
agricultural land, and the northern and western parts are largely
publicly owned forested land.  There is, however, a substantial area of
privately owned agricultural land in the Peace River Country of
northwestern Alberta.  There are also large areas in National
Parks along the Rocky Mountains and in the far north of the province.
The province calls the area of mostly publicly owned forested land
(excluding those National Parks) the Green Area, and the area of
mostly privately owned agricultural land the White Area.  About 90%
of the harvested timber volume in Alberta comes from the Green Area.

For administrative purposes, the Green Area is divided into Forest
Management Units (FMUs).  The province enters into long term
agreements with forest products companies giving the companies the
rights to timber harvest and the responsibility to manage the forest
sustainably.  These agreements are called Forest Management
Agreements, and the areas under agreement are called Forest Management
Agreement Areas (FMAAs).  For the most part, the boundaries of the
FMAAs are coincident with FMU boundaries, with the exception of
the Alberta-Pacific FMAA in eastern Alberta which comprises 12 FMUs.  Maps
of the FMUs @AlbertaFMU and the FMAAs @AlbertaFMA can be found on the
Government of Alberta website.  The FMU and FMAA boundaries are
important to this study as they are used to assign average piece size
and utilization standards.  We discuss piece sizes and utilization
standards later.

Our focus for this study is the forest near Hinton, Edson, and
Whitecourt in west-central Alberta.  These locations were chosen as
they host several forest products companies and are located near
forest areas with substantial logging activity.  We assume that a
residual forest biomass processing plant would be located near an
existing forest products mill as the forest road network has been
developed to supply those mills. Locating a forest residue processing
plant near a forest products mill would also allow for convenient
access to mill residues.

@@

Altalis manages much of the spatial data created by or for the
Government of Alberta.  We use several coverages from their base
features database product @Altalis2021: boundaries for the province,
the Green and White Areas, forest management units (FMUs), forest
management agreement areas (FMAAs), Alberta Township Survey System
(ATS_ townships, and towns.

The ATS boundaries are particularly important to this study as all the
data used for this study are aggregated or disaggregated to the
township level.  ATS provides the basis for legal land descriptions in
Alberta.  The typical township is roughly square (9.778 km by 9.716
km, 9 500 ha) and is indexed by township (1--126), range (1--30), and
meridian (4--6).  The townships immediately east of Alberta's western
boundary, and immediately east of the 5^th^ and 6^th^ meridians
(corresponding to 114°W and 118°W longitude, respectively) are partial
townships and therefore deviate from the roughly square shape.  ATS is
described in detail by the Alberta Land Surveyor's Association
@ALSA2021.  There are 7 237 townships in Alberta, allowing for spatial
resolution appropriate for regional or provincial level analyses.

@@

Alberta Geospatial Services is operated by the Alberta Ministry of
Environment and Parks @Alberta2021a.  We obtained road access
coverages from there.  A road access coverage is also available from
Altalis but it does not have as detailed information on road class.

@@

The National Forest Information System (NFIS) provides a set
of spatial data layers representing forest attributes at a 250 m
resolution for all of Canada @Beaudoin2017.  We used the merchantable
volume, stand age, and species composition attributes from these
data layers.

Another product of NFIS is a set of spatial datasets derived from
satellite data @NFIS2021a.  The dataset used for this study shows the
location of timber harvest for each of the years 1985-2015.  The
dataset is described in detail in Ref.  @Hermosilla2016.  It consists
of a raster for Canada at 30 m resolution indicating the year of
harvest for each cell.

@@

Ref.  @Ung2013 describes the tree taper models developed for Canada.
~~A valuable contribution related to this is the WWW-based wood volume
calculator that uses those models @NRCan2015.~~  We use these taper
models to determine merchantable volumes and the amount of residual
biomass available from cutblocks in each of the townships in each
year.

@@

The National Forestry Database @CCFM2021 was used to summarize the
timber harvest volume and harvest area from Alberta for the years
1990-2015. This information is used to adjust the volumes obtained
from the 250 m resolution inventory to better reflect average harvest
volumes observed in Alberta.

@@

The Alberta government requires FMA holders to develop a forest
management plan for the FMAA.  The forest management plans were used
to obtain information on utilization standards and average piece size of
harvested trees for the FMAA @Alberta2021.

@@

The boundaries of Canada's terrestrial ecozones can be found on
Agriculture and Agri-Food Canada's website @Canada2017.  Ecozone is an
input to the individual tree biomass calculator @NFIS2021b.

@@

The primary objective of this paper was to quantify the spatial and
temporal variability of the costs associated with recovering the
residual biomass left on forest cutblocks following timber harvest.  A
driving assumption for this model was that residue recovery is
opportunistic.  Loggers make the decision about what and where to
harvest, and the biomass recovery occurs at some time following timber
harvest.  The decision to harvest is made independently of any
knowledge of the value of residue as biomass.  The biomass
~~gleaners~~ harvesters
provide a service to the loggers as the residue would be piled and 
burned if it was not recovered as biomass.

Every township is  associated with a cycle time (and cost)
required to transport harvest residue from the township to each of our
3 plant locations, a measure of residual biomass for each of the 26
years in the study period. For each plant location and year, the model
examines all the townships in the province that have a positive
residual volume, and begins the creation of cumulative residual
biomass curves starting with the lowest cycle time first.  These lists
of cycle time and cumulative biomass provide the data for the marginal
cycle time and cost curves.

@@

The provincial road network was divided into four classes: primary
highways, other paved roads, two-lane gravel roads, and one-lane
gravel or dirt roads.  Loaded and unloaded travel speeds were assumed
for each road class, according to Table 1. These travel speeds are
based on those assumed by FPInnovations for the Dawson Creek Timber
Supply Area in the neighbouring province of British Columbia
@Blackburn2020.

The harmonic mean of loaded and unloaded speeds was calculated using Eq. 1.

$$
	\bar{s_h}=\left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}
$$

where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.  Use
of harmonic mean allows us to calculate the correct cycle time: the
time required to travel from the mill, to the collection site at the
cutblock, and back to the mill.  This provides the basis for our
estimates of haul cost.

Figure 2 illustrates the road network in Alberta and estimates of
cycle time based on the average of loaded and unloaded speeds
from Table 1. The road network in the White Area of the province is
very dense and largely arranged as a grid related to the Alberta
Township Survey System.  Most of the roads in the White Area are
public.  The road network in the Green Area is less dense and
many of the roads are built and maintained by resource industries
(*e.g.* forestry and petroleum & natural gas).  The pattern of the
road network is not as regular as in the White Area.

[ Figure 2 about here]

One-way haul times based on the harmonic mean travel speed were
generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
~~created by Raffler~~ @Raffler2018, specifically its iso-area as
interpolation (from point) algorithm.  In our case, we used the
algorithm to determine the minimum one-way travel time (using the
harmonic mean of loaded and unloaded speeds) along the road network
from each of our centres to each cell of a provincial level raster at
a 200 m resolution.  Off-road travel was assumed to be at 5 km/h.  The
Dijkstra algorithm @Dijkstra1959 is the basis for travel time
minimization.  Cycle times were calculated by multiplying the one-way
haul time by 2, and adding 1 hour for loading and unloading.

@@

@@

All measures of biomass used in this study are reported in SI units on
a dry matter basis:  1 Mg is therefore equivalent to 1 oven dry metric 
ton.

We assume that the biomass available for collection at a cutblock
includes the tops, branches, and foliage left at a roadside landing
after processing by a stroke delimber.  The delimber is assumed to top
the log at the top diameter specified in the utilization standard
relevant to the harvest area.
Our focus in this study is on the spatial and temporal variation in
biomass collection costs based on residue availability associated with
cycle times.  Following Ref. @Roser2013,
<!-- the pre-piling cost was set -->
<!-- to `r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution -->
<!-- cost (@Blackburn2020a) was set to `r sprintf("%5.2f",comminution_cost)` -->
<!-- \$·Mg^-1^ , and other costs (roads, -->
<!-- supervision, overhead, maintenance, and compliance) were set  -->
<!-- to `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .   -->
the cost of recovering residues from a roadside stroke-delimber , excluding
transport, is `r sprintf("%5.2f",constant_cost)` \$·Mg^-1^.
